---
title: "Both in One — 03a. Text and Emotion Analysis"
author: "Julieta Arco"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

---

### Overview ###

This notebook covers the preparation and analysis of the textual and emotional data collected during the *Both in One* painting process.

**Input files** (located in "here::here("Data")":

- `BiO_form.csv` — personal state variables per stage (exported from Jotform)
- `BiO_image.csv` — image metrics per stage (exported from Google Colab)

**Analysis steps:**

1. Data loading and preparation
2. Text preprocessing and word frequency analysis
3. Emotion analysis (NRC Emotion Lexicon)
4. Emotion word clouds by category

The combined dataset produced here (`BiO_data.csv`) serves as the input for `03b_Statistical_Analysis.Rmd`.

---

# 1. Data Preparation

```{r, warning=FALSE, message=FALSE}
# Load libraries
library(here)
library(dplyr)
library(readr)
library(lubridate)
library(tidyr)
library(stringr)
library(stringi)
library(tidytext)
library(stopwords)
library(ggplot2)
library(wordcloud)
library(scales)
library(syuzhet)
library(purrr)
```

```{r, warning=FALSE, message=FALSE}
# Load data
# Update file names accordingly
cat("Working directory:", here::here(), "\n")
image_data    <- read.csv(here::here("Data", "BiO_image.csv"))
form_data <- read.csv(here::here("Data", "BiO_form.csv"))
```


```{r}
# Sort by painting date (oldest to newest) and assign Stage number
form_df <- form_data %>%
  mutate(Date = mdy(Date)) %>%        
  arrange(Date) %>%                      # oldest stage first
  mutate(Stage = paste0("S", row_number())) 

```

```{r}
# Free-text reflective notes per stage

notes_data <- form_df %>%
  dplyr::select(
    Stage,
    Body_Notes        = "Body...Energy.Notes",
    Environment_Notes = "Environment.Notes",
    Work_Notes        = "Work.Notes",
    Social_Notes      = "Social.Notes",
    Intimacy_Notes    = "Intimacy.Notes",
    Emotion_Notes     = "Emotion.Notes",
    Leisure_Notes     = "Leisure.Notes",
    Spiritual_Notes   = "Spiritual.Notes",
    Growth_Notes      = "Growth.Notes",
    Creative_Notes    = "Creative.Notes"
  )

# Export data
write.csv(notes_data,
          file      = here::here("Data", "BiO_notes.csv"),
          row.names = FALSE)
```

```{r}
# Numerical personal state variables per stage
personal_data <- form_df %>%
  dplyr::select(
    Stage,
    Energy                  = "Energy.Level",
    Day_Structure           = "How.structured.or.chaotic.was.your.day.",
    Finance_Impact          = "To.what.extent.did.work.or.financial.matters.influence.you.today.",
    Social_Support          = "How.emotionally.supported.or.connected.did.you.feel.",
    Sexuality               = "How.connected.did.you.feel.with.your.intimacy.or.sexuality.",
    Emotional_Balance       = "How.balanced.or.grounded.did.you.feel.emotionally.",
    Mental_Focus            = "How.mentally.clear.or.focused.were.you.",
    Emotion_Intensity       = "How.intense.were.your.emotions.today.",
    Leisure_Time            = "How.much.time.did.you.dedicate.to.leisure.or.rest.",
    Spiritual_Connection    = "How.connected.did.you.feel.to.your.inner.purpose.or.spirituality.",
    Personal_Growth         = "How.aligned.did.you.feel.with.your.personal.values.or.growth.",
    Emotion_In_Painting     = "How.emotionally.present.were.you.while.painting.",
    Creative_Flow           = "How.strongly.did.you.feel.in.a.state.of.creative.flow.",
    Artistic_Satisfaction   = "How.satisfied.are.you.with.your.artistic.output.today.",
    Creative_Intention      = "How.clear.was.your.creative.intention.or.direction."
  )

# Export data
write.csv(personal_data,
          file      = here::here("Data", "BiO_personal.csv"),
          row.names = FALSE)
```


```{r}
# Combine all note columns into a single text column per stage
notes_data$all_Notes <- notes_data %>%
  dplyr::select(ends_with("_Notes")) %>%
  apply(1, function(x) paste(na.omit(x), collapse = " "))
```

---

# 2. Word Frequency Analysis

```{r, warning=FALSE, message=FALSE}
# Load libraries
library(stopwords)
library(tidytext)
library(stringi)
library(stringr)
```

```{r}
# Preprocess and tokenize
cleaned_words <- notes_data %>%
  dplyr::select(Stage = 1, all_Notes) %>%
  unnest_tokens(word, all_Notes) %>%
  filter(
    !word %in% stopwords::stopwords("es"),   # remove Spanish stopwords
    str_detect(word, "^[a-záéíóúñü]+$")      # keep only alphabetic words
  ) %>%
  mutate(word = stri_trans_general(word, "Latin-ASCII"))  # normalize accents
```

## 2.1 Word Frequency by Stage

```{r, warning=FALSE, message=FALSE}
# Load library
library(wordcloud)
```

```{r, fig.width=5, fig.height=5}
my_palette <- c("lightgrey", "darkblue", "violet", "purple",
                "purple3", "steelblue1", "magenta",
                "darkgrey", "blue")

stages <- unique(cleaned_words$Stage)

for (s in stages) {
  stage_data <- cleaned_words %>%
    filter(Stage == s) %>%
    count(word, sort = TRUE)

  wordcloud(words  = stage_data$word,
            freq   = stage_data$n,
            min.freq  = 1,
            max.words = 10,
            random.order = TRUE,
            scale  = c(2, 0.6),
            colors = my_palette)
  title(main = paste("Stage", s), cex.main = 1.5)
}
```

## 2.2 Overall Word Frequency

```{r, fig.width=6, fig.height=4}
word_freq <- cleaned_words %>%
  count(word, sort = TRUE)

# Bar chart — top 20 words
word_freq %>%
  slice_max(n, n = 20) %>%
  ggplot(aes(x = reorder(word, n), y = n)) +
  geom_col(fill = "magenta") +
  coord_flip() +
  labs(x = "Word", y = "Frequency") +
  theme_minimal()
```

```{r, fig.width=5, fig.height=5}
# Word cloud — overall
wordcloud(words  = word_freq$word,
          freq   = word_freq$n,
          max.words = 20,
          scale  = c(2, 0.6),
          colors = my_palette)
title(main = "Overall Word Frequency", cex.main = 1.5)
```

---

# 3. Emotion Analysis

Emotion scores are computed per stage using the NRC Emotion Lexicon via the `syuzhet` package.
The lexicon classifies words into eight basic emotions (Joy, Trust, Anticipation, Surprise, Fear, Sadness, Disgust, Anger) and two general sentiments (Positive, Negative).

```{r}
# Load library
library(syuzhet)
```

```{r}
# Compute NRC emotion scores per stage
emotion_scores <- get_nrc_sentiment(notes_data$all_Notes, language = "spanish")

# Rename columns
notes_emotions <- emotion_scores %>%
  rename(
    Anger        = anger,
    Anticipation = anticipation,
    Disgust      = disgust,
    Fear         = fear,
    Joy          = joy,
    Sadness      = sadness,
    Surprise     = surprise,
    Trust        = trust,
    Negative     = negative,
    Positive     = positive
  )

# Set display order
emotion_order <- c("Joy", "Trust", "Anticipation", "Surprise",
                   "Fear", "Sadness", "Disgust", "Anger",
                   "Positive", "Negative")

notes_emotions <- notes_emotions %>%
  dplyr::select(all_of(emotion_order))

# Combine with stage data
all_data <- bind_cols(notes_data, notes_emotions)
```

## 3.1 Overall Emotion Profile

```{r, fig.width=6, fig.height=4}
# Set colors
my_colors <- c(
  "Joy"          = "magenta",
  "Trust"        = "magenta3",
  "Anticipation" = "magenta4",
  "Surprise"     = "purple4",
  "Fear"         = "steelblue1",
  "Sadness"      = "steelblue3",
  "Disgust"      = "blue",
  "Anger"        = "darkblue",
  "Positive"     = "grey",
  "Negative"     = "grey1"
)

emotion_totals <- colSums(notes_emotions)

# Plot barplot
par(mar = c(8, 4, 4, 2))
barplot(emotion_totals,
        las       = 2,
        col       = adjustcolor(my_colors, alpha.f = 0.6),
        cex.names = 0.7,
        cex.axis  = 0.7,
        border    = NA,
        xlab      = "Emotion",
        ylab      = "Word Frequency",
        ylim      = c(0, max(emotion_totals) * 1.1),
        mgp       = c(3, 0.7, 0))
```

## 3.2 Emotion Trajectory Across Stages

```{r, fig.width=8, fig.height=4}
# Data format
emotion_long <- all_data %>%
  dplyr::select(Stage, all_of(emotion_order)) %>%
  pivot_longer(-Stage, names_to = "Emotion", values_to = "Score")

emotion_long$Emotion <- factor(emotion_long$Emotion, levels = emotion_order)

# Order stages numerically
stage_order <- emotion_long %>%
  distinct(Stage) %>%
  mutate(stage_num = as.numeric(gsub("[^0-9]", "", Stage))) %>%
  arrange(stage_num) %>%
  pull(Stage)

emotion_long$Stage <- factor(emotion_long$Stage, levels = stage_order)

# Plotter Emotions across Stages
ggplot(emotion_long, aes(x = Stage, y = Score, fill = Emotion)) +
  geom_col(position = "stack", alpha = 0.6) +
  scale_fill_manual(values = my_colors) +
  scale_x_discrete(drop = FALSE) +
  labs(x = "Stage", y = "Word Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## 3.3 Emotion Word Clouds

Word clouds per emotion category, showing only words associated with each emotion.

```{r, fig.width=5, fig.height=5}
# Match words to emotions
unique_words  <- unique(cleaned_words$word)
emotion_word  <- get_nrc_sentiment(unique_words, language = "spanish")

emotion_word <- emotion_word %>%
  mutate(word = unique_words) %>%
  relocate(word, .before = 1)

emotion_word_long <- emotion_word %>%
  pivot_longer(cols = -word, names_to = "emotion", values_to = "score") %>%
  filter(score > 0)

emotions_list <- unique(emotion_word_long$emotion)

for (emo in emotions_list) {
  emo_words <- emotion_word_long %>%
    filter(emotion == emo) %>%
    group_by(word) %>%
    summarise(freq = sum(score), .groups = "drop") %>%
    arrange(desc(freq))

# Plot words per emotion  
  wordcloud(words        = emo_words$word,
            freq         = emo_words$freq,
            min.freq     = 1,
            max.words    = 20,
            scale        = c(2, 0.6),
            colors       = my_palette)
  title(main = paste("Emotion:", emo), cex.main = 1.5)
}
```

---

# 4. Combined Dataset

```{r}
# Join image metrics, personal variables, and emotion scores by Stage
emotion_data <- all_data %>%
  dplyr::select(Stage, all_of(emotion_order))

df_stage <- image_data %>%
  inner_join(personal_data, by = "Stage") %>%
  inner_join(emotion_data,  by = "Stage")

glimpse(df_stage)
```

```{r}
# Export dataset for use in 03b_Statistical_Analysis.Rmd
write.csv(df_stage,
          file      = here::here("Data", "BiO_data.csv"),
          row.names = FALSE)
```

---

*Both in One — Julieta Arco*
