---
title: "Both in One — 03b. Statistical Analysis"
author: "Julieta Arco"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

---

### Overview ###

This notebook covers the statistical analysis of the *Both in One* dataset, exploring relationships between image metrics, personal state variables, and emotion scores.

**Input file** (produced by `03a_Text_and_Emotion_Analysis.Rmd`):

- `BiO_data.csv` — merged dataset with image metrics, personal variables, and emotion scores per stage

**Analysis steps:**

1. Exploratory Data Analysis (EDA) — distributions and trajectories
2. Correlation analysis
3. Hierarchical clustering
4. Principal Component Analysis (PCA)
5. Partial Least Squares Regression (PLSR)

---

# 1. Data Preparation

```{r, warning=FALSE, message=FALSE}
# Load libraries
library(here)
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(scales)
library(corrplot)
library(Hmisc)
library(pheatmap)
library(factoextra)
library(pls)
```

```{r, warning=FALSE, message=FALSE}
# Load full dataset
cat("Working directory:", here::here(), "\n")
df_stage <- read.csv(here::here("Data", "BiO_data.csv"))

# Separate variable groups for grouped plots
image_vars    <- c("Stroke_Density_Pct", "Texture_LBP_SD",
                   "Fractal_Dimension", "Mean_Hue", "Hue_SD")
emotion_order <- c("Joy", "Trust", "Anticipation", "Surprise",
                   "Fear", "Sadness", "Disgust", "Anger",
                   "Positive", "Negative")

# Personal variable names: all columns not in Stage, image_vars, or emotion_order
personal_vars <- setdiff(names(df_stage),
                         c("Stage", image_vars, emotion_order))

# All numeric variables for multivariate analysis
df_variables <- df_stage %>%
  dplyr::select(-Stage)

# Z-score scaling
df_variables_z <- as.data.frame(scale(df_variables, center = TRUE, scale = TRUE))
```

---

# 2. Exploratory Data Analysis

## 2.1 Distributions

### Image Metrics

```{r, fig.width=10, fig.height=3}
# Define color palette
color_img <- colorRampPalette(c("purple", "magenta", "magenta4", "blue", "darkblue"))(5)

# Plot boxplot
df_stage %>%
  dplyr::select(Stage, all_of(image_vars)) %>%
  pivot_longer(-Stage, names_to = "Metric", values_to = "Value") %>%
  ggplot(aes(x = "", y = Value, fill = Metric)) +
  geom_boxplot(width = 0.4, alpha = 0.6,
               outlier.shape = 16, outlier.size = 2,
               color = "black") +
  scale_fill_manual(values = color_img) +
  facet_wrap(~Metric, scales = "free", nrow = 1) +
  labs(x = NULL, y = "Value") +
  theme_minimal() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position    = "none",
    strip.text = element_text(face = "bold", size = 10)
  )
```

### Personal Variables

```{r, fig.width=10, fig.height=4}
# Define color palette
color_pers <- colorRampPalette(c("magenta", "grey", "darkblue"))(length(personal_vars))

# Plot boxplot
df_stage %>%
  dplyr::select(Stage, all_of(personal_vars)) %>%
  pivot_longer(-Stage, names_to = "Variable", values_to = "Score") %>%
  ggplot(aes(x = Variable, y = Score, fill = Variable)) +
  geom_boxplot(alpha = 0.6,
               outlier.shape = 16, outlier.size = 2) +
  scale_fill_manual(values = color_pers) +
  scale_y_continuous(breaks = function(x) seq(0, max(x), by = 2)) +
  labs(x = "Variable", y = "Score") +
  theme_minimal() +
  theme(
    axis.text.x    = element_text(angle = 90, hjust = 1),
    legend.position = "none"
  )
```

### Emotion Scores

```{r, fig.width=8, fig.height=4}
# Define color for each emotion
my_colors <- c(
  "Joy"          = "magenta",
  "Trust"        = "magenta3",
  "Anticipation" = "magenta4",
  "Surprise"     = "purple4",
  "Fear"         = "steelblue1",
  "Sadness"      = "steelblue3",
  "Disgust"      = "blue",
  "Anger"        = "darkblue",
  "Positive"     = "grey",
  "Negative"     = "grey1"
)

# Plot boxplot
df_stage %>%
  dplyr::select(all_of(emotion_order)) %>%
  pivot_longer(everything(), names_to = "Emotion", values_to = "Score") %>%
  mutate(Emotion = factor(Emotion, levels = emotion_order)) %>%
  ggplot(aes(x = Emotion, y = Score, fill = Emotion)) +
  geom_boxplot(alpha = 0.6,
               outlier.shape = 16, outlier.size = 2) +
  scale_fill_manual(values = my_colors) +
  scale_y_continuous(breaks = function(x) seq(0, max(x), by = 2)) +
  labs(x = "Emotion", y = "Score") +
  theme_minimal() +
  theme(
    axis.text.x    = element_text(angle = 90, hjust = 1),
    legend.position = "none"
  )
```

## 2.2 Trajectories Across Stages

Metric and variable values plotted across painting stages to visualize temporal evolution.

### Image Metrics

```{r, fig.width=3, fig.height=7}
# Select variables and change format
image_long <- df_stage %>%
  dplyr::select(Stage, all_of(image_vars)) %>%
  pivot_longer(-Stage, names_to = "Metric", values_to = "Value")

image_long$Stage <- factor(
  image_long$Stage,
  levels = paste0("S", sort(as.numeric(gsub("S", "", unique(image_long$Stage)))))
)

# Plot metrics
metrics_plot <- ggplot(image_long, aes(x = Stage, y = Value, group = Metric)) +
  geom_line(color  = "darkblue",  linewidth = 0.7) +
  geom_point(color = "magenta4",  size = 1.5) +
  facet_wrap(~Metric, scales = "free_y", ncol = 1) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 4)) +
  labs(x = "Stage", y = "Value") +
  theme_minimal(base_size = 11) +
  theme(
    panel.background  = element_rect(fill = "white"),
    panel.spacing     = unit(0.5, "lines"),
    panel.grid.major  = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.minor  = element_blank(),
    strip.text        = element_text(face = "bold"),
    axis.text.x       = element_text(angle = 90, hjust = 1)
  )

metrics_plot
```

### Personal Variables

```{r, fig.width=3, fig.height=15}
# Change format
personal_long <- df_stage %>%
  dplyr::select(Stage, all_of(personal_vars)) %>%
  pivot_longer(-Stage, names_to = "Variable", values_to = "Value")

personal_long$Stage <- factor(
  personal_long$Stage,
  levels = paste0("S", sort(as.numeric(gsub("S", "", unique(personal_long$Stage)))))
)

# Plot personal variables
personal_plot <- ggplot(personal_long, aes(x = Stage, y = Value, group = Variable)) +
  geom_line(color = "darkblue", linewidth = 0.7) +
  geom_point(color = "magenta4", size = 1.5) +
  facet_wrap(~Variable, scales = "free_y", ncol = 1) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) +
  labs(x = "Stage", y = "Value") +
  theme_minimal(base_size = 10) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.spacing    = unit(0.8, "lines"),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold", size = 9),
    axis.text.x      = element_text(angle = 90, hjust = 1, size = 7),
    axis.text.y      = element_text(size = 7),
    axis.title       = element_text(size = 9)
  )

personal_plot
```

---

# 3. Correlation Analysis

*Pearson correlations* between all variables (image metrics, personal states, emotion scores).
Significant correlations (p < 0.05) are marked with asterisks.

```{r, warning=FALSE, message=FALSE}
# Load libraries
library(Hmisc)
library(corrplot)
```

```{r, fig.width=6, fig.height=6}
# Pearson's Correlations
cor_results <- rcorr(as.matrix(df_variables), type = "pearson")
cor_matrix  <- cor_results$r
p_matrix    <- cor_results$P
diag(p_matrix) <- 0

# Define color gradient
color_gradient <- colorRampPalette(c("darkblue", "white", "magenta4"))(100)

# Plot correlations
corrplot::corrplot(cor_matrix,
                   method    = "color",
                   type      = "upper",
                   col       = color_gradient,
                   p.mat     = p_matrix,
                   sig.level = 0.05,
                   insig     = "label_sig",
                   pch.cex   = 1,
                   pch.col   = "black",
                   tl.col    = "black",
                   tl.cex    = 0.6,
                   cl.cex    = 0.6,
                   mar       = c(0, 0, 2, 0))
```

---

# 4. Hierarchical Clustering

Variables are clustered based on Euclidean distance (complete linkage) after z-score scaling, revealing groups of co-varying variables across stages.

```{r, warning=FALSE, message=FALSE}
# Load library
library(pheatmap)
```

```{r, fig.width=8, fig.height=5}
# Define color gradient
color_heat <- colorRampPalette(c("magenta4", "white", "darkblue"))(15)

# Plot heatmap
pheatmap(df_variables_z,
         cluster_rows         = FALSE,
         cluster_cols         = TRUE,
         clustering_distance_cols = "euclidean",
         clustering_method    = "complete",
         scale                = "none",
         fontsize_row         = 8,
         fontsize_col         = 10,
         color                = color_heat)
```

```{r, fig.width=8, fig.height=5}
# Dendrogram of variables
dist_vars <- dist(t(df_variables_z), method = "euclidean")
hc_vars   <- hclust(dist_vars, method = "complete")

# Plot dendrogram
plot(hc_vars, xlab = "", sub = "", cex = 0.8)
rect.hclust(hc_vars, k = 4,
            border = c("magenta", "magenta4", "blue", "darkblue"))
```

---

# 5. Principal Component Analysis (PCA)

*PCA* reduces the dimensionality of the dataset and identifies the main axes of covariation across all variables.

```{r, warning=FALSE, message=FALSE}
# Load library
library(factoextra)
```

```{r }
# Define color gradient
color_pca <- colorRampPalette(c("darkblue", "white", "magenta4"))(30)

# PCA Analysis
pca_result <- prcomp(df_variables, center = TRUE, scale. = TRUE)
summary(pca_result)
```

```{r, fig.width=6, fig.height=5, warning=FALSE}
# Plot PCA with variable contributions
fviz_pca_var(pca_result,
             col.var      = "contrib",
             gradient.cols = color_pca,
             labelsize    = 2,
             repel        = TRUE,
             title        = "PCA — Variable Contributions")
```

---

# 6. Partial Least Squares Regression (PLSR)

*PLSR* identifies latent components that maximize the shared covariance between image metrics (predictors, X) and state/emotion variables (responses, Y). Model performance is evaluated using leave-one-out cross-validation (LOO-CV).

**Conceptual model:**

> Y [State or Emotion] ~ X [Stroke Density, Texture LBP SD, Fractal Dimension, Mean Hue, Hue SD]

```{r, warning=FALSE, message=FALSE}
# Load library
library(pls)
```

```{r}
# Define predictors and responses
# Update response variable names to match the dataset
predictors_df <- df_variables %>%
  dplyr::select(all_of(image_vars))

responses_df <- df_variables %>%
  dplyr::select(all_of(c(personal_vars, emotion_order)))

X <- as.matrix(predictors_df)
Y <- as.matrix(responses_df)
```

## 6.1 Component Selection

```{r, fig.width=6, fig.height=4}
# Fit PLSR with LOO cross-validation across all possible components
pls_model <- plsr(Y ~ X,
                  data       = df_variables,
                  scale      = TRUE,
                  validation = "LOO")

summary(pls_model)

# Compute mean RMSEP across all responses to determine optimal components
rmseps_matrix <- RMSEP(pls_model)$val[1, , ]  # dims: components x responses
rmseps_means  <- colMeans(rmseps_matrix[-1, ]) # remove intercept row

optimal_ncomp <- which.min(rmseps_means)
cat("Optimal number of components:", optimal_ncomp, "\n")

# Plot mean RMSEP
plot(seq_along(rmseps_means), rmseps_means,
     type = "n",
     xlab = "Number of Components",
     ylab = "Mean RMSEP",
     main = "Mean RMSEP Across All Responses")
grid()
lines(seq_along(rmseps_means), rmseps_means,
      type = "b", pch = 19, col = "darkblue", lwd = 2, cex = 1.2)
abline(h   = min(rmseps_means), col = "magenta4", lty = 2)
abline(v   = optimal_ncomp,    col = "steelblue1", lty = 2)
```

## 6.2 Final Model

```{r}
# Fit final model with optimal number of components
pls_final <- plsr(Y ~ X,
                  ncomp      = optimal_ncomp,
                  data       = df_variables,
                  scale      = TRUE,
                  validation = "LOO",
                  method     = "kernelpls")

summary(pls_final)
```

## 6.3 Predicted vs. Observed

```{r, fig.width=10, fig.height=8}
# Cross-validated predictions
predicted_vals <- predict(pls_final, ncomp = optimal_ncomp, type = "response")
predicted_vals <- predicted_vals[, , 1]
pred_df        <- as.data.frame(predicted_vals)
colnames(pred_df) <- colnames(responses_df)

# Plot predicted vs observed for each response variable
n_vars <- ncol(responses_df)
n_cols <- 4
n_rows <- ceiling(n_vars / n_cols)

par(mfrow = c(n_rows, n_cols), mar = c(3, 3, 2, 1))

for (var in colnames(responses_df)) {
  plot(responses_df[[var]], pred_df[[var]],
       xlab = "Observed", ylab = "Predicted",
       main = var, pch = 16, col = "steelblue1")
  abline(0, 1, col = "magenta4", lty = 2)
}
```

## 6.4 Model Performance (R²)

```{r}
# R² per response variable (cross-validated)
r2_values <- sapply(colnames(responses_df), function(var) {
  obs  <- responses_df[[var]]
  pred <- pred_df[[var]]
  ss_res <- sum((obs - pred)^2)
  ss_tot <- sum((obs - mean(obs))^2)
  if (ss_tot == 0) return(NA)
  return(1 - ss_res / ss_tot)
})

r2_df <- data.frame(
  Variable = names(r2_values),
  R2       = round(r2_values, 3)
) %>%
  arrange(desc(R2))

print(r2_df)
```

```{r, fig.width=7, fig.height=5}
# Bar chart of R² values
r2_df %>%
  filter(!is.na(R2)) %>%
  ggplot(aes(x = reorder(Variable, R2), y = R2)) +
  geom_col(fill = "steelblue1", alpha = 0.8) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.3) +
  coord_flip() +
  labs(x = "Response Variable", y = expression(R^2),
       title = "PLSR Cross-Validated R² per Response Variable") +
  theme_minimal() +
  theme(plot.title = element_text(size = 11))
```

---

*Both in One — Julieta Arco*
